\begin{abstract}
Archaic hominin introgression has shaped the genome of modern humans,
yet existing detection methods quantify introgression through archaic variant density,
discarding the haplotype match-length signal that coalescent theory predicts to
differ $\sim$11-fold between introgressed and non-introgressed segments.
Here we introduce \textbf{ArchaicPainter} (\AP{}), a three-state
(AMH / Neanderthal / Denisovan) Hidden Markov Model that extends the
Li \& Stephens haplotype copying framework to archaic reference genomes,
explicitly encoding match-length information in coalescent-theoretic emission
and transition probabilities.
In 50 independent simulation replicates with known ground truth,
\AP{} achieves segment-level F1 $= 0.480 \pm 0.095$ (AUPRC $= 0.577$)
versus F1 $= 0.009 \pm 0.003$ for a Poisson density baseline
($56\times$ improvement; Wilcoxon $p = 8.9 \times 10^{-16}$),
quantifying the discriminative value of haplotype structure
over density-counting alone.
Ablation analysis reveals that the gain derives from a bidirectional emission model
exploiting both matches and mismatches as archaic-state evidence;
a positive-only emission mode --- appropriate when the available archaic reference
diverges from the true introgressing population --- is provided for real-data use.
Applied to chromosome~21 from 1000~Genomes Project Phase~3 data,
\AP{} detects $2.28\% \pm 1.22\%$ Neanderthal ancestry in Europeans (CEU)
and $2.09\% \pm 0.77\%$ in East Asians (CHB)
(median segment length $\sim$90--113~kb), consistent with published genome-wide
estimates and complementary to IBDmix's high-sensitivity short-tract detection.
A three-state extension achieves proof-of-concept simultaneous Neanderthal and
Denisovan source attribution (NEA F1 $= 0.260$, DEN F1 $= 0.188$),
and near-linear computational scaling (empirical exponent 1.21, $0.11$~s per
haplotype at $n = 1{,}000$) supports population-scale deployment.
Functional annotation of detected chr21 regions overlaps with 7 of 16 published
adaptive introgression candidates and a complete interferon receptor gene cluster
(\textit{IFNAR1}, \textit{IFNAR2}, \textit{IFNGR2}, \textit{IL10RB}) in an
unsupervised manner, consistent with adaptive retention of archaic antiviral
immunity haplotypes in Europeans.
\AP{} is released as open-source software with reproducible pipelines
for community reuse.
\end{abstract}
