\section*{Materials and Methods}

\subsection*{The \AP{} Hidden Markov Model}

\paragraph{Overview.}
\AP{} infers archaic introgression at haplotype resolution using a three-state
local ancestry HMM whose transition probabilities are parameterized by
the Li \& Stephens (2003) tract-length formulation~\citep{li2003modeling},
adapted here so that hidden states represent \emph{ancestry categories}
(AMH, NEA, DEN) rather than individual reference haplotypes.
The core insight is coalescent-theoretic: a modern human haplotype segment that traces
ancestry to an archaic hominin via introgression coalesces with the archaic reference
at time $t_\mathrm{admixture} \approx 1{,}724$ generations ($\sim$50 kya), whereas a
non-introgressed segment coalesces at the modern human--archaic split time
$t_\mathrm{split} \approx 19{,}000$ generations ($\sim$550 kya).
Under a Poisson recombination model, the expected tract length scales as $1/t$ Morgans,
yielding an 11-fold difference in expected haplotype match lengths
($\approx$58 kb for introgressed versus $\approx$5 kb for non-introgressed haplotypes
at a typical recombination rate of 1~cM/Mb).
\AP{} exploits this \emph{match-length signal} by parameterizing the
archaic-state persistence probability using $t_\mathrm{admixture}$
(encoding the $\approx$58-kb introgressed tract length),
while the AMH background rate is set by the expected inter-introgression
spacing ($d_\mathrm{AMH} \approx 2.9$~Mb) rather than by $t_\mathrm{split}$
directly; the 11-fold ratio provides the conceptual motivation for the
length-based approach rather than being operationalized as two explicit
scale parameters in the model (Figure~1).
Methods such as HMMix~\citep{skov2018detecting}, DAIseg~\citep{planche2024daiseg},
and IBDmix~\citep{chen2020identifying} model archaic introgression via Poisson
counts of archaic-specific alleles within sliding windows, an architecture that
discards all haplotype linkage disequilibrium structure.
The tract-length transition formulation is analogous to modern local ancestry
inference (HAPMIX~\citep{price2009sensitive};
MOSAIC~\citep{salter2019mosaic}; ChromoPainter~\citep{lawson2012inference}),
here adapted to archaic reference genomes.

\paragraph{Hidden states.}
\AP{} assigns each phased SNP position on a query haplotype to one of three
mutually exclusive hidden states: $S_\mathrm{AMH}$ (modern human ancestry),
$S_\mathrm{NEA}$ (Neanderthal-introgressed), and $S_\mathrm{DEN}$
(Denisovan-introgressed).
For analyses involving a single archaic reference (all chromosome~21 real-data
experiments), the DEN state is inactive and the model reduces to a two-state HMM.
Initial state probabilities are set to $\pi_\mathrm{NEA} = 0.02$ and
$\pi_\mathrm{AMH} = 0.98$, consistent with published introgression fractions;
results are robust to this prior over long sequences.

\paragraph{Emission probabilities.}
At each biallelic SNP site $i$, the query haplotype carries allele $a_i \in \{0, 1\}$.
Let $f_i^\mathrm{YRI}$ denote the derived allele frequency estimated from Yoruban
(YRI) haplotypes in the 1000~Genomes Project Phase~3 panel~\citep{genomes20151000}.
Under $S_\mathrm{AMH}$:
\begin{equation}
  P(a_i \mid S_\mathrm{AMH}) = f_i^\mathrm{YRI} \cdot [a_i=1] + (1 - f_i^\mathrm{YRI}) \cdot [a_i=0].
  \label{eq:emiss_amh}
\end{equation}

Under $S_\mathrm{NEA}$, two emission models are used depending on whether the
archaic reference genome matches the true introgressing population.

\emph{Standard bidirectional emission (simulation).}
When the reference is drawn from the same simulated population as the true donor,
mismatches are informative and the standard Li \& Stephens emission is used:
\begin{equation}
  P(a_i \mid S_\mathrm{NEA}, \text{sim.}) =
  (1-\theta)[a_i = g_i^\mathrm{arch}] + \theta[a_i \neq g_i^\mathrm{arch}],
  \label{eq:emiss_nea_std}
\end{equation}
where $\theta = 0.01$ is the per-site mismatch probability between the query
haplotype and the archaic reference (a composite parameter capturing
sequencing error, ancient DNA damage, and residual reference--donor
divergence in simulation; not the germline mutation rate $\mu \approx 10^{-8}$),
and $g_i^\mathrm{arch}$ is the archaic reference allele at site $i$.
On the 50-replicate benchmark this formulation achieves F1 $= 0.480 \pm 0.095$.

\emph{Positive-only log-score (real data).}
Because Vindija was not the direct donor of introgressed segments in living
Europeans or East Asians, divergence between Vindija and the true donor population
means mismatches at a given site may reflect inter-population variation rather
than absence of introgression.
We therefore down-weight mismatch evidence via the following log-scoring function
for real-data analyses:
\begin{equation}
  \ell(a_i \mid S_\mathrm{NEA}, \text{real}) =
  \begin{cases}
    \log(1 - \theta) & \text{if } a_i \text{ matches } g_i^\mathrm{arch}, \\
    \log\!\tfrac{1}{2}        & \text{otherwise.}
  \end{cases}
  \label{eq:emiss_nea_pos}
\end{equation}
\textbf{Mathematical note (important limitation).}
These log-scores do \emph{not} define a normalized probability distribution:
$e^{\log(1-\theta)} + e^{\log(1/2)} = (1-\theta) + 1/2 > 1$ for $\theta < 1/2$.
Because the excess mass $(1-\theta) + 1/2 - 1 \approx 0.49$ constitutes a
per-site additive bonus in log-space for the NEA state, the Viterbi path is
\emph{biased toward the NEA state} relative to a properly normalized model:
a run of $L$ sites in the NEA state accumulates an extra log-score of
$\approx 0.49 L$ compared to a normalized model.
This systematic bias inflates the effective length prior for NEA segments and
shifts the detection threshold; the magnitude depends on the ratio of NEA
to AMH per-site emission sums.
Posterior probabilities derived from this model are approximate ranking
scores rather than calibrated probabilities and should be interpreted
as heuristic confidence measures.
The $r$-mixture model below provides a properly normalized alternative for
future work.

A properly normalized alternative is the $r$-mixture model:
\begin{align}
  P(a_i = \text{match} \mid S_\mathrm{NEA}, r) &= r(1-\theta) + (1-r)/2, \nonumber \\
  P(a_i = \text{mismatch} \mid S_\mathrm{NEA}, r) &= r\theta + (1-r)/2,
  \label{eq:r_mixture}
\end{align}
which sums to 1 for all $r \in [0,1]$, reduces to the standard bidirectional
model at $r=1$, and to a fully uninformative emission ($P=1/2$ for both outcomes)
at $r=0$.
The log-score in Equation~\eqref{eq:emiss_nea_pos} approximates the $r$-mixture in
the regime where mismatch evidence is strongly suppressed (small $r$) while match
evidence is retained from Equation~\eqref{eq:emiss_nea_std}.
A full treatment with estimated $r$ from population-genetic divergence parameters
is a natural extension; we adopt the log-score formulation here as a computationally
convenient heuristic for this demonstration.

In practice, setting the mismatch log-score to $\log(1/2)$ substantially reduces
the log-odds penalty for mismatches relative to the standard model
($\log\theta = \log 0.01 \approx -4.6$ vs.\ $\log(1/2) \approx -0.69$),
making the NEA state largely agnostic to disagreements.
Note that mismatch is not strictly neutral relative to AMH:
the AMH emission uses $f^\mathrm{YRI}_i$, so the log-odds at a mismatch site equals
$\log(1/2) - \log(1 - f^\mathrm{YRI})$, which is mildly negative (weak evidence
against NEA) for typical derived-allele frequencies.
Critically, the standard and positive-only models are \emph{not} interchangeable:
on simulation, the positive-only model achieves F1 $= 0.154$ versus $0.480$ for
the bidirectional model, because mismatches carry essential discriminative information
when the reference exactly matches the introgression source population.
The positive-only model is used exclusively for real-data analyses; the
bidirectional model is used for all simulation benchmarks.

\paragraph{Transition probabilities.}
Let $d_i$ (Morgans) be the genetic distance between sites $i$ and $i+1$
from a genetic map.
Under the Haldane recombination model~\citep{price2009sensitive}:
\begin{equation}
  P(S_{i+1} = s \mid S_i = s) = e^{-d_i \cdot t_s},
  \label{eq:trans}
\end{equation}
where $d_s = 1/t_s^\mathrm{eff}$ (Morgans) is the expected tract length for state $s$.
For the two archaic states, $t_s^\mathrm{eff}$ is the admixture time:
$t_\mathrm{NEA}^\mathrm{eff} = 1{,}724$ generations ($d_\mathrm{NEA} \approx 58$~kb)
and $t_\mathrm{DEN}^\mathrm{eff} = 1{,}552$ generations ($d_\mathrm{DEN} \approx 64$~kb).
For the AMH background state, which spans ${\sim}98\%$ of the genome, the
expected block length between introgression events is
$d_\mathrm{AMH} = 1/(\pi_\mathrm{arch} \cdot t_\mathrm{NEA}^\mathrm{eff})
\approx 1/(0.020 \times 1{,}724) \approx 2.9$~Mb,
where $\pi_\mathrm{arch}$ is the total archaic prior.
This parameterisation ensures AMH is the stable background with transitions to
archaic states occurring at the rate consistent with the introgression fraction,
and is numerically equivalent to setting
$P(\text{stay AMH}) = e^{-d_i \cdot \pi_\mathrm{arch} \cdot t_\mathrm{NEA}^\mathrm{eff}}$.
Off-diagonal probabilities are distributed proportional to state priors.
Genetic distances are computed by linear interpolation of the GRCh38
sex-averaged recombination map~\citep{halldorsson2019characterizing}
(HapMap-II format, UCSC Genome Browser).

\paragraph{Viterbi decoding and post-processing.}
State sequences are inferred by the Viterbi algorithm in log-space.
Consecutive archaic-state positions are merged into segments.
For real-data analyses, gaps $\leq 10$~kb between adjacent segments are closed
(reducing fragmentation from sparse Vindija sites) and merged segments shorter
than 10~kb are discarded.
These thresholds were selected by grid search on the simulation benchmark
to maximize F1 and held fixed for all real-data runs (sensitivity:
Supplementary Table~S3).

\subsection*{Simulation Experiments}

Coalescent simulations used \texttt{msprime}~v1.3.3~\citep{baumdicker2022efficient}
(Python~3.12).
The single-source benchmark demography modeled YRI (reference), CEU (query),
and NEA (Neanderthal reference) populations.
A population split at $t_\mathrm{OOA} = 2{,}000$ generations separated YRI and CEU,
followed by a 2\% admixture pulse from NEA into CEU at $t_\mathrm{admixture} = 1{,}724$
generations via \texttt{msprime.MassMigration}.
The NEA--modern split was at $t_\mathrm{split} = 19{,}000$ generations~\citep{prufer2014complete};
all $N_e = 10{,}000$ diploid.

The 50-replicate benchmark used seeds 42--91 (5~Mb each, 20 CEU query haplotypes,
2-haplotype NEA reference).
Ground-truth intervals were extracted from all ARG migration records with times
within $\pm 80\%$ of $t_\mathrm{admixture}$.

The two-source proof-of-concept used a two-archaic demography with
MassMigration events at $t_\mathrm{NEA} = 1{,}724$ gen (2\%) and $t_\mathrm{DEN} = 1{,}000$
gen (1\%), and population splits at $t_\mathrm{split\_NEA} = 19{,}000$ gen and
$t_\mathrm{split\_DEN} = 25{,}000$ gen.
A simulated Denisovan reference was used to ensure reference--source correspondence.
Twenty replicates were run (seeds 100--119, 5~Mb, 20 query haplotypes).

\subsection*{Evaluation Metrics}

\emph{F1}: true positive = predicted segment overlaps $\geq 1$~bp of a ground-truth segment
(lenient, following Sankararaman et al.~\citeyear{sankararaman2014genomic};
stricter 50\% and 80\% reciprocal-overlap results in Supplementary Table~S4).
\emph{AUPRC}: precision-recall curve integrated over Viterbi log-score thresholds
(valid as a ranking metric for simulation, where the bidirectional emission is a
proper probability model; for real-data positive-only scores the curve measures
ranking quality but not probability calibration).
\emph{Attribution accuracy}: fraction of predicted NEA (or DEN) segments overlapping
a same-type ground-truth segment; \emph{cross-contamination}: fraction overlapping
the opposite-type ground truth.
\emph{Bootstrap CIs}: percentile bootstrap, 2{,}000 resamples (SciPy~1.14.1~\citep{virtanen2020scipy}).
\emph{Wilcoxon test}: one-sided signed-rank test for pairwise method comparisons.

\subsection*{Real-Data Analysis}

\paragraph{Data.}
Phased chromosome~21 VCFs were downloaded from the 1000~Genomes Phase~3 GRCh38
release (ftp.1000genomes.ebi.ac.uk)~\citep{genomes20151000}.
We analyzed 20 CEU and 20 CHB individuals (40 phased haplotypes each)
across chr21:10,000,000--46,709,983 (GRCh38 chromosome end; $\approx$36.7~Mb).
YRI allele frequencies were estimated from 20 Yoruban individuals (40 haplotypes)
in the same release; frequencies below 0.025 were floored at 0.025 to prevent
emission instability from singletons.

\paragraph{Vindija reference.}
The Vindija~33.19 diploid VCF~\citep{prufer2017high} was downloaded from the
MPI EVAN repository, converted from GRCh37 to GRCh38 via CrossMap~v0.7.0~\citep{zhao2014crossmap},
and merged with the 1000GP VCF using allele-aware orientation correction
(swapped REF/ALT codes were inverted before computing match scores).
Heterozygous Vindija positions (where the two alleles differ) are treated as
uninformative in the archaic emission ($P = 0.5$, equivalent to log-odds~0),
since the Viterbi path cannot determine which Vindija haplotype, if any, is
ancestral to the introgressed segment; sites with missing Vindija genotypes
are handled identically.
After intersection and quality filtering, 1{,}036 (CEU) and 906 (CHB) informative
SNPs remained ($\approx$27 SNPs/Mb), substantially fewer than in simulation
($\approx$320 SNPs/Mb) due to the limited callable regions of the Vindija genome.
The median inter-SNP distance ($\approx$37~kb) retains sensitivity to tracts
$\geq 50$~kb but reduces boundary resolution relative to simulation.

\paragraph{Ancestry fraction.}
Per-haplotype Neanderthal ancestry fraction is the proportion of the 36.7-Mb analysis window
covered by NEA-state Viterbi calls.
Population estimates are mean $\pm$ SD across the 40 haplotypes.

\subsection*{Two-Source Attribution}
\label{sec:attribution}

The three-state HMM was evaluated on the two-source simulation (20 replicates,
5~Mb, 2\% NEA + 1\% DEN) described above, using a simulated Denisovan reference
matched to the true donor.
Real-data Denisovan attribution using the Altai genome is deferred to future work
pending integration of a phased, high-coverage Denisovan haplotype reference for
GRCh38 that satisfies the phased-VCF input requirements of the pipeline;
published GRCh38-aligned Denisovan variant sets currently available (e.g., UCSC
\texttt{denisovan.hg38.filt.vcf.gz}) are unphased and would require additional
phasing or imputation steps before use with \AP{}.

\subsection*{Functional Annotation}

Introgressed segment coordinates were intersected with UCSC knownGene
annotations (hg38, downloaded February~2026)~\citep{kent2002human};
gene symbols were mapped via the UCSC kgXref table.
A transcript was considered overlapping at $\geq 1$~bp.
Named genes were cross-referenced against a curated list of chr21 adaptive
introgression loci (Supplementary Table~S2) compiled from Vernot \& Akey~\citeyear{vernot2014resurrecting},
Sankararaman et al.~\citeyear{sankararaman2014genomic},
Vernot et al.~\citeyear{vernot2016excavating},
Browning et al.~\citeyear{browning2018ancestry},
and Dannemann \& Kelso~\citeyear{dannemann2017contribution}.

\subsection*{Scalability Analysis}

Runtime was benchmarked with query sizes $n = 10$ to $n = 1{,}000$ haplotypes
(5-Mb sequences, seed 42, single CPU thread, Intel Xeon Platinum 8358P, 2.60~GHz).
The empirical scaling exponent was estimated by log-log regression of total
runtime against $n$.
Theoretical per-haplotype complexity is $O(L \times K^2)$ where $L$ = sites,
$K$ = states.

\subsection*{Software and Reproducibility}

\AP{} is implemented in Python~3.12; code, simulation scripts, preprocessing
pipelines, and pinned dependencies (\texttt{environment.yml}) are provided
in Supplementary Scripts~S1--S5 and at \url{https://github.com/EvoClaw/ArchaicPainter}.
